---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

The following is the code I used to load the data to a variable called "data". 
I also wanted to see how the data looks like, so I used the head function.

```{r}
data<- read.csv("activity.csv")
head(data)
```


## What is mean total number of steps taken per day?

### Calculate for the total number of steps taken per day.

I need to find out the average steps per day, and arrange them in a table using the aggregate function. Note that the problem asks for the TOTAL (sum) of the steps taken per day. I also used the xtable function to make the table look more presentable.

```{r echo=TRUE,results='asis'}

library(xtable)
stepsperday<-aggregate(steps~date, data=data, rm.na=TRUE, FUN=sum)
print(xtable(stepsperday), type = "html")

```

### Make a histogram of the number of steps taken per day.
For this plot, I will use the ggplot plotting system, taking the dates in the x-axis and then the average number of steps in the y-axis.

```{r echo=TRUE, fig.height=6, fig.width=12}

library(ggplot2)
meanplot<-ggplot(data,aes(x=factor(date),y=steps))+ stat_summary(fun.y="mean", geom="bar")
meanplot + xlab("Date")+ylab("Average Steps")+ggtitle("Average Number of Steps per Day") + theme(axis.text.x=element_text(angle=60, hjust=1))

```

### Calculate and report the mean and median of the total number of steps taken per day.

For this, I will also use the aggregate function. First, the mean:

```{r echo=TRUE, results='asis'}

ave<-aggregate(steps~date, data=data, rm.na=TRUE, FUN=mean)
print(xtable(ave), type = "html")

```

Next, the median:

```{r echo=TRUE, results='asis'}

medi<-aggregate(steps~date, data=data, rm.na=TRUE, FUN=median)
print(xtable(medi), type = "html")

```


## What is the average daily activity pattern?

### Make a time series plot of the time intervals.

For this, I will use the base plotting system.

``` {r echo=TRUE,fig.height=6, fig.width=12}

inter<-aggregate(steps~interval,data=data,rm.na=TRUE,FUN=mean)
plot(inter,type="l",xlab="Time Intervals",ylab="Average Number of Steps",main="Average Number of Steps in Time Intervals")

```

### Which 5-minute interval contains the maximum number of steps?

Since we already have a data frame of the aggregated average steps with time intervals ("inter"), I will simply get the max from that.

```{r echo=TRUE}

sortinter<-inter[order(-inter$steps),]
sortinter[1,]

```



## Imputing missing values

### Counting the number of NA rows

For this, I will use the is.na function, with colSums so we know which columns have the corresponding number of NAs.

```{r echo=TRUE}

colSums(is.na(data))

```


### Filling in the NA rows

I used the na.omit function:

```{r echo=TRUE}

newdata<-data
newdata$steps[is.na(newdata$steps)]<-mean(na.omit(data$steps))


```

### Looking at the difference

```{r echo=TRUE, fig.height=6, fig.width=12}

newmeans<-ggplot(newdata,aes(x=factor(date),y=steps))+ stat_summary(fun.y="mean", geom="bar")
newmeans + xlab("Date")+ylab("NEW Average Steps")+ggtitle("Average Number of Steps per Day") + theme(axis.text.x=element_text(angle=60, hjust=1))

```

NEW means:

```{r echo=TRUE, results='asis'}

newave<-aggregate(steps~date, data=newdata, rm.na=TRUE, FUN=mean)
print(xtable(newave), type = "html")

```

NEW medians:

```{r echo=TRUE, results='asis'}

newmedi<-aggregate(steps~date, data=newdata, rm.na=TRUE, FUN=median)
print(xtable(newmedi), type = "html")

```

When comparing these results, there IS a difference, but only a slight one in a few days (like the first day, for example).

## Are there differences in activity patterns between weekdays and weekends?

### Adding a variable indicating whether the date is a weekend or a weekday (in column "wDay")

``` {r echo=TRUE}

newdata$date <- as.Date(newdata$date)
weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
newdata$wDay <- factor((weekdays(newdata$date) %in% weekdays1), 
         levels=c(FALSE, TRUE), labels=c('weekend', 'weekday')) 

```

Look at the new data here:

``` {r echo=TRUE, results='asis'}

newdata$date <- data$date
print(xtable(head(newdata)),type="html")

```

### Comparing the plots of weekday and weekend based on time intervals

``` {r echo=TRUE,fig.height=12, fig.width=12}

daydata<-newdata[newdata$wDay=="weekday",]
enddata<-newdata[newdata$wDay=="weekend",]

par(mfrow = c(2, 1))
dayplot<-aggregate(steps~interval,data=daydata,rm.na=TRUE,FUN=mean)
endplot<-aggregate(steps~interval,data=enddata,rm.na=TRUE,FUN=mean)
plot(dayplot,type="l",xlab="Time Intervals",ylab="Average Number of Steps",main="Average Number of Steps in Time Intervals on WEEKDAYS")
plot(endplot,type="l",xlab="Time Intervals",ylab="Average Number of Steps",main="Average Number of Steps in Time Intervals on WEEKENDS")


```
